Behavioural Analysis of Independent Value-Based Learning in Non-cooperative Games

### 2025-06-05 paper/paper.tex
Update paper.tex

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
%\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled
\usepackage{svg}

\usepackage[numbers,sort&compress]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{[}{]}{,}{n}{,}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
\makeatletter% @ becomes a letter
\def\NAT@def@citea{\def\@citea{\NAT@separator}}% Suppress spaces between citations using natbib.sty
\makeatother% @ becomes a symbol again

@@

\begin{abstract}
Multi-agent reinforcement learning has received a lot of updates in cooperative games recently. However, research in general non-cooperative games is currently lagging behind. Value-based algorithms such as Q-learning have been one of the most widely studied methods in reinforcement learning, and their independent variations for multi-agent environments are popular approaches due to their simplicity and good scalability. Some research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios. In this paper, we discuss the feasibility of these algorithms in non-cooperative settings. As a result, we explain the conditions a game must satisfy for the algorithms to work. We further propose the game Food Chain. The game simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our formal analysis. Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games.
\end{abstract}

@@

\subsection{Stochastic Game}
A game is formally defined by a tuple:
$G = \langle I,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
\begin{itemize}
    \item $N = \{1, ..., n\}$: the set of agents, numbered from 1 to $n$.
    \item $S$: the game’s state space.
    \item $S_0 \subset S$: the set of possible initial states.
    \item $S_t \subset S$: the set of terminal states.
    \item $\{A_1,\ldots,A_n\}$: one action space for each agent. Let $a_i \in A_i$ be an action of agent $i$, and $a^n=\langle a_1,\ldots,a_n \rangle \in A^n$ denote a joint action in the joint action space.
    \item $\{R_1,\ldots,R_n\}$: one reward function for each agent. $R_i:S \times A^n \times S \rightarrow \mathbb{R}$.
    \item $T$: the state transition function. $T:S \times A^n \times S \rightarrow [0,1]$. It defines the probability of the game transitions from one state to another given the joint action. It must satisfy the axiom of probability:
    \begin{equation}
        \forall s \in S,a^n \in A^n: \sum_{s' \in S}{T(s,a^n,s')}=1
        \label{eq:state_transition}
    \end{equation}
\end{itemize}
Starting from a randomly chosen $s_0 \in S_0$, at each time step $t \in \mathbb{N}$ with the current state $s_t$, each agent observes the current state $s_t$ and chooses an action forming a joint action $a^n=\langle a_1,\ldots,a_n \rangle$. The game then moves into the next state $s_{t+1}$, randomly sampled from the distribution P and a reward $R_i$ is assigned to each agent. If $s_t \in S_t$, the game ends. Otherwise, the game continues in the next time step.

>>>>

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
%\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled
\usepackage{svg}
%\usepackage{biblatex}
\usepackage[numbers,sort&compress]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{[}{]}{,}{n}{,}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
\makeatletter% @ becomes a letter
\def\NAT@def@citea{\def\@citea{\NAT@separator}}% Suppress spaces between citations using natbib.sty
\makeatother% @ becomes a symbol again

@@

\begin{abstract}
Multi-agent reinforcement learning has received a lot of updates in cooperative games recently. However, research in general non-cooperative games is currently lagging behind. Independent value-based algorithms such as independent Q-learning have been one of the most widely studied methods in the field due to their simplicity and scalability. Some research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios. In this paper, we discuss the feasibility of these algorithms in non-cooperative settings. As a result, we explain the conditions a game must satisfy for the algorithms to work. We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our formal analysis. Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
\end{abstract}

@@

\subsection{Stochastic Game}
A game is formally defined by a tuple:
$G = \langle N,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
\begin{itemize}
    \item $N = \{1, ..., n\}$: the set of agents, numbered from 1 to $n$.
    \item $S$: the game’s state space.
    \item $S_0 \subset S$: the set of possible initial states.
    \item $S_t \subset S$: the set of terminal states.
    \item $\{A_1,\ldots,A_n\}$: one action space for each agent. Let $a_i \in A_i$ be an action of agent $i$, and $a^n=\langle a_1,\ldots,a_n \rangle \in A^n$ denote a joint action in the joint action space.
    \item $\{R_1,\ldots,R_n\}$: one reward function for each agent. $R_i:S \times A^n \times S \rightarrow \mathbb{R}$.
    \item $T$: the state transition function. $T:S \times A^n \times S \rightarrow [0,1]$. It defines the probability of the game transitions from one state to another given the joint action. It must satisfy the axiom of probability:
    \begin{equation}
        \forall s \in S,a^n \in A^n: \sum_{s' \in S}{T(s,a^n,s')}=1
        \label{eq:state_transition}
    \end{equation}
\end{itemize}
Starting from a randomly chosen $s_0 \in S_0$, at each time step $t \in \mathbb{N}$ with the current state $s_t$, each agent observes the current state $s_t$ and chooses an action forming a joint action $a^n=\langle a_1,\ldots,a_n \rangle$. The game then moves into the next state $s_{t+1}$, randomly sampled from the distribution P and a reward $R_i$ is assigned to each agent. If $s_t \in S_t$, the game ends. Otherwise, the game continues in the next time step.
